---
title: "MID -SEM MATHEMATICLA MODELLING"
author: "NEEMA NDANU"
date: "2024-06-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QUESTION 1
### Given Information:
-> Parameters of Interest: Contact rate (20), Transmission probability (0.07), Infectiousness period (12), latency period (21)
-> Disease Dynamic Parameters and Initial Values: Exposed (15,000), Susceptible (20,989),Infected (11), Recovered (0), Dead (0)
-> States: Susceptible, Exposed, Infected, Recovered, Hospitalized, Dead
-> Time Period for Monitoring: 180 days
### Present the Schematic Framework illustrate the compartments and interactions between age-groups. Remember that susceptible children experience the force of infection, while susceptible adults become infected at a constant rate. For this case assume that:[5 marks]-> Children are 30% of the entire population.-> Children make 20 contacts per day on average, with 12 contacts being with other children.−> Adults make 20 contacts per day on average, with 15 contacts being with other adults.


```{r}
# Load the necessary packages
library(deSolve)
library(tidyr)
library(tidyverse)
library(DiagrammeR)
library(ggplot2)
library(reshape2)
library(gridExtra)

# Define population parameters
initial_susceptible <- 20989
initial_infected <- 11
initial_recovered <- 0
initial_exposed <- 15000
initial_hospitalized <-0
initial_dead <- 0 
population_size <- initial_susceptible + initial_infected + initial_exposed + initial_hospitalized + initial_recovered + initial_dead
S_child <- 0.3 * population_size
S_adult <- population_size - S_child

# Define disease parameters
infectious_period <- 12
latency_period <- 21
transmission_prob <- 0.07
recovery_rate <- 1/infectious_period
simulation_period <- 180 # in days

# Contact rate parametres
contact_rate <- 20 
contact_rate_child <- 20
contact_rate_adult <- 20
contact_adult_adult <- 15
contact_adult_child <- contact_rate_adult- contact_adult_adult  
contact_child_child <- 12
contact_child_adult <- contact_rate_child - contact_child_child

# For child alone
S_child <- 0.3 * initial_susceptible
E_child <- 0.3 * 15000
I_child <- 0.3 * 11
R_child <- 0
H_child <- 0
D_child <- 0

# For adults alone:
S_adult <- (1-0.3) * initial_susceptible
E_adult <- (1-0.3) * initial_exposed
I_adult <- (1-0.3) * initial_infected 
R_adult <- 0
H_adult <- 0
D_adult <- 0

# Schematic framework for the whole poputlation
DiagrammeR::grViz("
digraph SEIR {
  rankdir = LR;
  node [shape = circle, fontsize = 20, penwidth = 2.0]
  S [label = 'S']
  E [label = 'E']
  I [label = 'I']
  R [label = 'R']
  D [label = 'D']
  
  edge [arrowhead = vee]
  S -> E [label = 'β']
  E -> I [label = 'k']
  I -> H [label = 'Ih']
  I -> R [label = 'γ']
  I -> D [label = 'α']
  H -> D [label = 'μh']
  H -> R [label = 'μμ']
  
}
")

# Schematic framework for children and adults in the population
DiagrammeR::grViz("
digraph SEIR {
  rankdir = LR;
  
  node [shape = circle, fontsize = 20, penwidth = 2.0]
  
  # Nodes for children
  S_child [label = 'S_child']
  E_child [label = 'E_child']
  I_child [label = 'I_child']
  R_child [label = 'R_child']
  D_child [label = 'D_child']
  H_child [label = 'H_child']
  
  # Nodes for adults
  S_adult [label = 'S_adult']
  E_adult [label = 'E_adult']
  I_adult [label = 'I_adult']
  R_adult [label = 'R_adult']
  D_adult [label = 'D_adult']
  H_adult [label = 'H_adult']
  
  edge [arrowhead = vee]
  
  # Transitions for children
  S_child -> E_child [label = 'β_child']
  E_child -> I_child [label = 'k_child']
  I_child -> H_child [label = 'Ih_child']
  I_child -> R_child [label = 'γ_child']
  I_child -> D_child [label = 'α_child']
  H_child -> D_child [label = 'μh_child']
  H_child -> R_child [label = 'μμ_child']
  
  # Transitions for adults
  S_adult -> E_adult [label = 'β_adult']
  E_adult -> I_adult [label = 'k_adult']
  I_adult -> H_adult [label = 'Ih_adult']
  I_adult -> R_adult [label = 'γ_adult']
  I_adult -> D_adult [label = 'α_adult']
  H_adult -> D_adult [label = 'μh_adult']
  H_adult -> R_adult [label = 'μμ_adult']
  
  # Interactions between children and adults
  I_child -> E_adult [label = 'β_ca', style=dotted]
  I_adult -> E_child [label = 'β_ac', style=dotted]
}
")



```

S(t) The number of susceptible individuals at a given time, t
E(t) The number of exposed individuals at a given time, t
I(t) The number of infected individuals at a given time, t
R(t) The number of recovered individuals at a given time, t
D(t) The number of dead individuals at a given time t

β  The exposure rate.
β_ca the rate at which children who are infected are exposed to the adults
β_cc the rate at which children who are infected are exposed to other children
β_aa the rate at which adults who are infected are exposed to the adults
β_ac the rate at which adults who are infected are exposed to the other adults
γ The rate at which an infected individuals recovered per unit time
k The rate of exposed individuals are infected
α The rate at which infected die
µµ The rate at which those who went to the hospital are not hospitalized
µh The rate at which those who went to the hospital and are hospitalized
Ih Th rate at which those are infected went to the hospital and are hospitalized, when less than one it shows limited bed capacity and grater than one more bed capacity or unlimited bed capacity.

## QUESTION 2
### Simulate the SEIRD Epidemic based on the provided parameters and initial values.[10 marks]−> Define the SEIRD model equations incorporating age-group-specific contact rates.−>Utilize the provided initial values to simulate the epidemic over a 180-day period.−> Incorporate hospitalization. take into consideration two scenarios unlimited and limited bed capacity.− For hospitalized individuals, specify a mortality rate due to critical illness.

```{r}
# Define the hospital parameters
death_rate <- 0.4
uu <- 0.9
uh <- 1 - uu

# Define the SEIRD model function
seird_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    # Total population sizes
    N_child <- S_c + E_c + I_c + R_c + H_c + D_c
    N_adult <- S_a + E_a + I_a + R_a + H_a + D_a
    
    # Force of infection for children and adults
    lambda_c <- beta_child * (I_c + I_a) / (N_child + N_adult)
    lambda_a <- beta_adult * (I_c + I_a) / (N_child + N_adult)
    
    # SEIR equations for children
    dS_c <- -lambda_c * S_c
    dE_c <- lambda_c * S_c - kappa_child * E_c
    dI_c <- kappa_child * E_c - (1 / infectious_period) * I_c - uh_child * I_c
    dR_c <- (1 / infectious_period) * I_c
    if (H_c < Ih_child) {
      # If hospital capacity is not full for children
      dH_c <- uh_child * (1 / infectious_period) * I_c - gamma_child * H_c
      dD_c <- death_rate * H_c
    } else {
      # If hospital capacity is full for children
      dH_c <- 0
      dD_c <- death_rate * I_c
    }
    
    # SEIR equations for adults
    dS_a <- -lambda_a * S_a
    dE_a <- lambda_a * S_a - kappa_adult * E_a
    dI_a <- kappa_adult * E_a - (1 / infectious_period) * I_a - uh_adult * I_a
    dR_a <- (1 / infectious_period) * I_a
    if (H_a < Ih_adult) {
      # If hospital capacity is not full for adults
      dH_a <- uh_adult * (1 / infectious_period) * I_a - gamma_adult * H_a
      dD_a <- death_rate * H_a
    } else {
      # If hospital capacity is full for adults
      dH_a <- 0
      dD_a <- death_rate * I_a
    }
    
    # Return the rate of change
    return(list(c(dS_c, dE_c, dI_c, dR_c, dH_c, dD_c,
                  dS_a, dE_a, dI_a, dR_a, dH_a, dD_a)))
  })
}


# Initial state values
initial_state_values <- c(
  S_c = S_child,
  E_c = E_child,
  I_c = I_child,
  R_c = R_child,
  H_c = H_child,
  D_c = D_child,
  S_a = S_adult,
  E_a = E_adult,
  I_a = I_adult,
  R_a = R_adult,
  H_a = H_adult,
  D_a = D_adult
)

# Parameters for unlimited hospital capacity
parameters_unlimited <- list(
  beta_child = contact_child_adult * transmission_prob,
  kappa_child = 1 / latency_period,
  gamma_child = 1 / infectious_period,
  uh_child = uh,
  alpha_child = death_rate,
  Ih_child = Inf,
  beta_adult = contact_adult_adult * transmission_prob,
  kappa_adult = 1 / latency_period,
  gamma_adult = 1 / infectious_period,
  uh_adult = uh,
  alpha_adult = death_rate,
  Ih_adult = Inf
)

# Parameters for limited hospital capacity
parameters_limited <- list(
  beta_child = contact_child_adult * transmission_prob,
  kappa_child = 1 / latency_period,
  gamma_child = 1 / infectious_period,
  uh_child = uh,
  alpha_child = death_rate,
  Ih_child = 200,
  beta_adult = contact_adult_adult * transmission_prob,
  kappa_adult = 1 / latency_period,
  gamma_adult = 1 / infectious_period,
  uh_adult = uh,
  alpha_adult = death_rate,
  Ih_adult = 300
)

# Define the time vector and simulation period
time <- seq(0, simulation_period, by = 1)

# Solve the ODE system for unlimited hospital capacity
out_unlimited <- ode(y = initial_state_values, times = time, func = seird_model, parms = c(parameters_unlimited,parameters_unlimited))

# Solve the ODE system for limited hospital capacity
out_limited <- ode(y = initial_state_values, times = time, func = seird_model, parms = c(parameters_unlimited,parameters_unlimited))

# Convert the output to data frames
output_df_unlimited <- as.data.frame(out_unlimited)
colnames(output_df_unlimited) <- c("time", "S_c", "E_c", "I_c", "R_c", "H_c", "D_c", "S_a", "E_a", "I_a", "R_a", "H_a", "D_a")

output_df_limited <- as.data.frame(out_limited)
colnames(output_df_limited) <- c("time", "S_c", "E_c", "I_c", "R_c", "H_c", "D_c", "S_a", "E_a", "I_a", "R_a", "H_a", "D_a")

# Combine results into tidy format for plotting
dat_unlimited <- output_df_unlimited %>%
  pivot_longer(cols = c(S_c, E_c, I_c, R_c, H_c, D_c, S_a, E_a, I_a, R_a, H_a, D_a),
               names_to = "state", values_to = "value")

dat_limited <- output_df_limited %>%
  pivot_longer(cols = c(S_c, E_c, I_c, R_c, H_c, D_c, S_a, E_a, I_a, R_a, H_a, D_a),
               names_to = "state", values_to = "value")


```

### −> Plot the trajectories of Susceptible, Exposed, Infected, Recovered, Hospitalized, and Dead sub-populations over time.
```{r}
# PLot the graphs
plot_children_unlimited <- ggplot(subset(dat_unlimited, grepl("_c", state)), aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIRD Children (Unlimited)",
       x = "Time",
       y = "Population",
       color = "State") +
  scale_color_manual(values = c("S_c" = "blue", "E_c" = "orange", "I_c" = "red", "R_c" = "green", "H_c" = "purple", "D_c" = "black")) +
  theme_minimal()

plot_children_limited <- ggplot(subset(dat_limited, grepl("_c", state)), aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIRD Children (Limited)",
       x = "Time",
       y = "Population",
       color = "State") +
  scale_color_manual(values = c("S_c" = "blue", "E_c" = "orange", "I_c" = "red", "R_c" = "green", "H_c" = "purple", "D_c" = "black")) +
  theme_minimal()

plot_adults_unlimited <- ggplot(subset(dat_unlimited, grepl("_a", state)), aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIRD Adults (Unlimited)",
       x = "Time",
       y = "Population",
       color = "State") +
  scale_color_manual(values = c("S_a" = "blue", "E_a" = "orange", "I_a" = "red", "R_a" = "green", "H_a" = "purple", "D_a" = "black")) +
  theme_minimal()

plot_adults_limited <- ggplot(subset(dat_limited, grepl("_a", state)), aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIRD Adults (Limited)",
       x = "Time",
       y = "Population",
       color = "State") +
  scale_color_manual(values = c("S_a" = "blue", "E_a" = "orange", "I_a" = "red", "R_a" = "green", "H_a" = "purple", "D_a" = "black")) +
  theme_minimal()

# Display the plots
plot_children_unlimited
plot_children_limited
plot_adults_unlimited
plot_adults_limited
grid.arrange(plot_children_unlimited, plot_children_limited, plot_adults_unlimited, plot_adults_limited,
             nrow = 2, ncol = 2)
```
We can conclude the following form the above graph that with unlimited hospital bed capacity the rate of death decrease over time much faster but with limited hospital bed capacity the rate of death increases very faster before they start to decrease. This is because unlimited bed capacity means that the infected will not be able to get the treatment on time taus making them to worsen over time. Also with unlimited bed capacity the number of recovered will increase over time as most people will receive the treatment on time but with limited the recovery rate will increase slowly not as fast as when we have unlimited bead capacity.Therefore we can avoid such unnecessary death through increasing the bed capacity by building new ward or temporary facilities,develop emergency healthcare to work more effectively and maybe use of better technology to handle the outbreaks ans reduce the amount of infected thus reducing the deaths.

### QUESTION 3
### Evaluate and Discuss Changes with Vaccination assuming a vaccine with 100% effectiveness against acquiring infections and analyze its impact on sub-populations.− Introduce vaccination into the model, varying vaccination percentages (10%, 20%, 30%,50%). [3 marks] 

```{r}
# Function to simulate SEIRD model with vaccination
simulate_vaccination <- function(vac_rate) {
  # Adjust initial susceptible population based on vaccination rate
  initial_values_vac <- initial_state_values
  initial_values_vac["S_c"] <- initial_values_vac["S_c"] * (1 - vac_rate)
  initial_values_vac["S_a"] <- initial_values_vac["S_a"] * (1 - vac_rate)
  
  # Define parameters with vaccination effect
  parameters_vac <- list(
    beta_child = contact_child_adult * transmission_prob * (1 - vac_rate),
    beta_adult = contact_adult_adult * transmission_prob * (1 - vac_rate),
    kappa_child = 1 / latency_period,
    gamma_child = 1 / infectious_period,
    uh_child = uh,
    alpha_child = death_rate,
    Ih_child = Inf,
    N_child = sum(initial_values_vac[c("S_c", "E_c", "I_c", "R_c", "H_c", "D_c")]),
    kappa_adult = 1 / latency_period,
    gamma_adult = 1 / infectious_period,
    uh_adult = uh,
    alpha_adult = death_rate,
    Ih_adult = Inf,
    N_adult = sum(initial_values_vac[c("S_a", "E_a", "I_a", "R_a", "H_a", "D_a")])
  )
  
  # Solve the ODE system with vaccination parameters
  out_vac <- ode(y = initial_values_vac, times = time, func = seird_model, parms = parameters_vac)
  out_vac <- as.data.frame(out_vac)
  
  # Add vaccination rate as a column
  out_vac$vaccination_rate <- vac_rate
  
  return(out_vac)
}

# Vaccination scenarios: 10%, 20%, 30%, 50%
vaccination_rates <- c(0.1, 0.2, 0.3, 0.5)
results <- lapply(vaccination_rates, simulate_vaccination)

# Combine results for plotting
names(results) <- paste0("Vaccination_", vaccination_rates * 100, "%")
results_df <- do.call(rbind, results)
results_df$vaccination_rate <- factor(results_df$vaccination_rate)

```

### − Simulate and observe changes in the trajectories of Susceptible, Exposed, Infected,Recovered, Hospitalized, and Dead sub-populations. [4 marks]

```{r}
# Plotting the results for each sub-population and vaccination rate
library(ggplot2)

# Children plots
plot_children <- ggplot(results_df, aes(x = time)) +
  geom_line(aes(y = S_c, color = "Susceptible")) +
  geom_line(aes(y = E_c, color = "Exposed")) +
  geom_line(aes(y = I_c, color = "Infected")) +
  geom_line(aes(y = R_c, color = "Recovered")) +
  geom_line(aes(y = H_c, color = "Hospitalized")) +
  geom_line(aes(y = D_c, color = "Dead")) +
  labs(title = "SEIRD Model with Vaccination - Children",
       x = "Time (days)",
       y = "Population",
       color = "State") +
  facet_wrap(~ vaccination_rate, scales = "free_y") +
  scale_color_manual(values = c("blue", "orange", "red", "green", "purple", "black")) +
  theme_minimal()

# Adults plots
plot_adults <- ggplot(results_df, aes(x = time)) +
  geom_line(aes(y = S_a, color = "Susceptible")) +
  geom_line(aes(y = E_a, color = "Exposed")) +
  geom_line(aes(y = I_a, color = "Infected")) +
  geom_line(aes(y = R_a, color = "Recovered")) +
  geom_line(aes(y = H_a, color = "Hospitalized")) +
  geom_line(aes(y = D_a, color = "Dead")) +
  labs(title = "SEIRD Model with Vaccination - Adults",
       x = "Time (days)",
       y = "Population",
       color = "State") +
  facet_wrap(~ vaccination_rate, scales = "free_y") +
  scale_color_manual(values = c("blue", "orange", "red", "green", "purple", "black")) +
  theme_minimal()

# Display plots
plot_children
plot_adults

```


### Discuss how different vaccination rates influence the epidemic dynamics, hospitalization  rates, and mortality rates
 The higher you increase the vaccine rate the faster the  decline in the susceptible population,since more individuals acquire immunity, leading to a steady decrease in the susceptible population for both adults and children. Also the exposed don't change much since they are not affected by the vaccines rates . There is there is a noticeable reduction in the number of infected individuals this highlights the effectiveness of vaccination in curbing disease transmission and lastly the recovered increases over time reflecting both natural recovery from infection and immunity they acquire form vacinnation.