---
title: "MATHEMATICAL MODELLLING ASS 2"
author: "NEEMA NDANU"
date: "2024-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## BACKGROUND:
### Kenya, like many countries around the world, has faced challenges in managing the COVID19 pandemic. For this hypothetical scenario, we’ll consider a situation where the government of Kenya is analyzing the impact of COVID-19 spread and the effectiveness of vaccination strategies. We also want to visualize the dynamics of the epidemic under the given parameters.Parameters:Population Size: 50 million individuals,Initial Infected Individuals: 500, Initial Recovered Individuals: 0, Duration of Infectiousness: 10 days, Transmission Rate: 0.3 (per day), Recovery Rate: 1/10 (0.1 per day), Simulation Period: 180 days

### Question 1: Model Framework (5 Marks):- Write an R code snippet to initialize the parameters and initial state values for the given COVID-19 scenario in Kenya.
```{r}
# Define the population parameters
N <- 50000000
initial_infected <-500
initial_recovered <-0
initial_susceptible <- N- initial_infected - initial_recovered


# Define the disease parameters
duration_infection <-10
transimssion_rate <-0.3
recovery_rate <-0.1

# Simulation period
simulation_days<-180

```

### EXPLANATION
The susceptible population is the number of people likely to be harmed or influenced and since we have not yet defined it we can calculate it using the existing parameters that is the infected and the recovered parameters. susceptible = Population size - initial Infected population - Initial recovered population. 

### Question 2: Model Implementation (5 Marks):- a)Implement the SIR model in R using the parameters defined in the scenario.Your implementation should define the model equations and solve them over a period of 180 days.Explain the role of each parameter in the SIR model and how it influences the dynamics of the epidemic.
```{r}
# Load necessary libraries
library(deSolve)
library(DiagrammeR)
library(tidyr)

# Define the SIR model
sir_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    
    # SIR equations
    dS <- -beta * S * I / N
    dI <- beta * S * I / N - gamma * I
    dR <- gamma * I
    
    # Return the rate of change
    list(c(dS, dI, dR))
  })
}

# Initial state values
initial_state_values <- c(S = initial_susceptible, I = initial_infected, R = initial_recovered)

# Parameters
parameters <- c(beta = transimssion_rate , gamma = recovery_rate, N = N)

# Time vector
time <- seq(0, simulation_days, by = 1)

# Solve the model
output  <- ode(y = initial_state_values, times = time, func = sir_model, parms = parameters)

# Convert the output to a data frame
output <- as.data.frame(output)
colnames(output) <- c("time", "Susceptible", "Infected", "Recovered")

# Display dimensions of the output data frame
dim(output)

# Display the first few rows of the output data frame
knitr::kable(head(output, 10))

# Convert the data to a tidy format
dat <- output %>% pivot_longer(cols = c(Susceptible, Infected, Recovered), names_to = "state", values_to = "value")

# Display the first few rows of the tidy data
knitr::kable(head(dat, 10))
# Create a diagram to visualize the SIR model
DiagrammeR::grViz("
digraph SIR {
  rankdir = LR;
  node [shape = circle, fontsize = 20, penwidth = 2.0]
  S [label = 'S']
  I [label = 'I']
  R [label = 'R']
  
  edge [arrowhead = vee]
  S -> I [label = 'β = 0.3']
  I -> R [label = 'γ = 0.1']
}
")


```

### Explanation: 
->The data frame show how the S, I an R populations will looks over the time period. The diagram is a graphical representation of the SIR model.

->SIR Model Equations: The SIR model divides the population into three compartments: Susceptible (S), Infected (I), and Recovered (R). The dynamics of the epidemic are described by three ordinary differential equations (ODEs), which represent the rate of change of each compartment over time.

dS/dt = -β * S * I / N: This equation represents the rate of change of susceptible individuals over time. I

dI/dt = β * S * I / N - γ * I: This equation represents the rate of change of infected individuals over time. 

dR/dt = γ * I: This equation represents the rate of change of recovered individuals over time. It describes how infected individuals recover and become immune to the disease.

->Parameter Explanation:

β (transmission rate): This parameter measures the average number of people one infected individual infects per day. A higher β leads to faster transmission of the disease.

γ (recovery rate): This parameter measures the proportion of infected individuals recovering per day. It represents the inverse of the duration of infectiousness. A higher γ means individuals recover faster.

N (total population size): This parameter represents the total size of the population. It influences the rate at which susceptible individuals become infected and affects the dynamics of the epidemic.

->Simulation Period:-This period allows us to observe how the epidemic evolves over time, from the initial outbreak to potential control measures such as vaccinations. or natural immunity development including the recovery.

### Question 3: Data Analysis and Visualization (5 Marks):-Create a line plot showing the number of susceptible, infected, recovered, and vaccinated individuals over time for the model implemented.Ensure your plot is well-labeled and includes a legend.

```{r}
library(ggplot2)
# Define the vaccine parameters and initial vaccination state
output$Vaccinated <- 0

# Convert the data to a tidy format
dat <- output %>% pivot_longer(cols = c(Susceptible, Infected, Recovered, Vaccinated), names_to = "state", values_to = "value")

# Display the first few rows of the tidy data
knitr::kable(head(dat, 10))

# Plot the results
ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SIR Model of COVID-19 Epidemic",
       x = "Days",
       y = "Number of Individuals",
       color = "State") +
  scale_color_manual(values = c("red","green","blue", "purple")) +
  theme_minimal()
```

### EXPLANATION

The line plot generated by the code illustrates the number of susceptible, infected, recovered, and vaccinated individuals over time based on the implemented model. From the graph, we can observe that as the number of recovered individuals increases, the numbers of susceptible and infected individuals decrease. The point where the curves for susceptible and recovered individuals intersect corresponds to the peak of the infected curve. Additionally, the number of vaccinated individuals remains constant throughout the simulation. This constancy is due to the fact that no vaccination rate was specified, implying that the model assumes an instantaneous vaccination process, where the entire susceptible population could be vaccinated simultaneously. However, since the vaccination rate is not defined, the number of vaccinated individuals stays at zero, indicating that the vaccination effort was not successful in this scenario.

### Question 4: Intervention (5 Marks) :-Assume a vaccination of 20% of the population. Discuss how the vaccination rate affects the epidemic curve.

```{r}
# Vaccination parameters
vaccination_rate <- 0.2
initial_vaccinated <- initial_susceptible * vaccination_rate

# Update initial state values
initial_susceptible_vaccinated <- initial_susceptible - initial_vaccinated

# Initial state values with vaccination
initial_state_values_vaccinated <- c(S = initial_susceptible_vaccinated, I = initial_infected, R = initial_recovered)

# Solve the model with vaccination
out_vaccinated <- ode(y = initial_state_values_vaccinated, times = time, func = sir_model, parms = parameters)

# Convert the output to a data frame
out_vaccinated <- as.data.frame(out_vaccinated)
colnames(out_vaccinated) <- c("time", "Susceptible", "Infected", "Recovered")

# Add the vaccinated column
out_vaccinated$Vaccinated <- initial_vaccinated

# Convert the data to a tidy format
dat_vaccinated <- out_vaccinated %>% pivot_longer(cols = c(Susceptible, Infected, Recovered, Vaccinated), names_to = "state", values_to = "value")

# Plot the results with vaccination
ggplot(dat_vaccinated, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SIR Model of COVID-19 Epidemic with 20% Vaccination",
       x = "Days",
       y = "Number of Individuals",
       color = "State") +
  scale_color_manual(values = c("red","green","blue", "purple")) +
  theme_minimal()

```

### EXPLANATION

Introducing a vaccination intervention of 20% significantly reduces the number of infected individuals. Vaccination decreases the number of susceptible individuals, slowing virus transmission and reducing the peak of the epidemic curve. The plot illustrates fewer susceptible and infected individuals and more recovered individuals compared to the previous scenario without vaccination. Initially, the number of vaccinated individuals increases from 0 to 10 because 20% of the susceptible population is vaccinated at the start. However, after this initial increase, the number of vaccinated individuals remains constant throughout the simulation. This is because the model assumes an instantaneous vaccination process and no further vaccinations are administered during the simulation. Overall, this intervention leads to a flatter and lower peak of infections, highlighting the effectiveness of vaccination in controlling the epidemic.

### B) What changes do you observe in the number of infected individuals when the vaccination rate is doubled?Explain the importance of the transmission rate (𝛽) and how public health interventions could reduce this rate
```{r}
# Define vaccination parameters
# Doubled vaccination rate
vaccination_rate_doubled <- 0.4 

# Calculate initial vaccinated individuals with doubled vaccination rate
initial_vaccinated_doubled <- initial_susceptible * vaccination_rate_doubled

# Update initial state values with doubled vaccination rate
initial_susceptible_vaccinated_doubled <- initial_susceptible - initial_vaccinated_doubled

# Define initial state values with doubled vaccination rate
initial_state_values_vaccinated_doubled <- c(S = initial_susceptible_vaccinated_doubled, I = initial_infected, R = initial_recovered)

# Solve the model with doubled vaccination rate
out_vaccinated_doubled <- ode(y = initial_state_values_vaccinated_doubled, times = time, func = sir_model, parms = parameters)

# Convert the output to a data frame
out_vaccinated_doubled <- as.data.frame(out_vaccinated_doubled)
colnames(out_vaccinated_doubled) <- c("time", "Susceptible", "Infected", "Recovered")

# Add the vaccinated column
out_vaccinated_doubled$Vaccinated <- initial_vaccinated_doubled

# Convert the data to a tidy format
dat_vaccinated_doubled <- out_vaccinated_doubled %>% pivot_longer(cols = c(Susceptible, Infected, Recovered, Vaccinated), names_to = "state", values_to = "value")

# Plot the results with doubled vaccination rate
ggplot(dat_vaccinated_doubled, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SIR Model of COVID-19 Epidemic with Doubled Vaccination",
       x = "Days",
       y = "Number of Individuals",
       color = "State") +
  scale_color_manual(values = c("red","green","blue", "purple")) +
  theme_minimal()

```

### EXPLANATION

When the vaccination rate is doubled to 40%, the number of infected individuals decreases from approximately 9 million to around 4 million. This reduction is due to the increased immunity in the population, which lowers the transmission rate of the virus. With the higher vaccination rate, the number of vaccinated individuals rises from 10 to 20 million. Despite this increase, the number of vaccinated individuals remains constant throughout the 180 days, assuming an instantaneous vaccination process where a certain number of people get vaccinated every day.

The transmission rate (𝛽) is crucial in determining how quickly and extensively the virus spreads. Public health interventions, such as vaccination campaigns and mask-wearing, are essential in reducing 𝛽. By vaccinating more individuals, these interventions effectively decrease the number of susceptible individuals, slowing down virus transmission and reducing overall infections.

Doubling the vaccination rate not only decreases the number of infected individuals but also underscores the importance of public health efforts in controlling transmission rates to effectively manage the epidemic.
