---
title: "GROUP 5 MATHEMATICAL MODELLING"
author: "NEEMA NDANU"
date: "2024-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mathematical model for measles disease with control on the susceptible and  exposed compartment
### INSTRUCTIONS
You are required to review the paper assigned to your group and distill out information based on the 
following:
− Focus and motivation of the paper
− Data Source
− Model input parameters
− Model input values (state or compartment values)
− Model equations used
− Exploratory Data Analysis output

```{r}
# Load the necessary packages
library(deSolve)
library(tidyverse)
library(DiagrammeR)
library(ggplot2)
library(gridExtra)

# Define parameters
population_size <- 1000
initial_susceptible <- 800
initial_infected <- 10
initial_recovered <- 0
initial_exposed <- 200
exposure_rate <- 0.09  # round to 2dp
recruitment_rate <- 0.037 # round to 2dp
infected_rate <- 0.14 # round to 2 dp
leaving_rate <- 0.01241 # per day
recovery_rate <- 0.125 #  per day
simulation_period <- 10 # in years

# Create a diagram to visualize the SEIR model
DiagrammeR::grViz("
digraph SEIR {
  rankdir = LR;
  node [shape = circle, fontsize = 20, penwidth = 2.0]
  S [label = 'S']
  E [label = 'E']
  I [label = 'I']
  R [label = 'R']
  N [label = 'N']
  
  edge [arrowhead = vee]
  S -> E [label = 'β = 0.09']
  E -> I [label = 'α = 0.14']
  I -> R [label = 'γ = 0.125']
  E -> R [label = 'σ']
  S -> R [label = 'v']
  N -> S [label = 'b']
}
")
```

S(t) The number of susceptible individuals at a given time, t
E(t) The number of exposed individuals at a given time, t
I(t) The number of infected individuals at a given time, t
R(t) The number of recovered individuals at a given time, t

b Recruitment rate ( by birth and/or immigrants)
v vaccination rate for the susceptible class who later got vaccinated
µ Leaving rate (by death and/or emigrants)
β The contact rate
γ The rate at which an infected individuals recovered per unit time
σ The rate of exposed individuals who have undergone testing and therapy
α The rate at which an exposed become infected

```{r}
# Define the SEIR model
seir_model <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    
    # SEIR equations
    dS <- b - beta * S * I - (v + mu) * S
    dE <- beta * S * I  - (v + alpha + sigma) * E
    dI <- alpha * E - (mu + gamma) * I
    dR <- v * S + gamma * I + sigma * E - mu * R
    
    # Return the rate of change
    return(list(c(dS, dE, dI, dR)))
  })
}

# Initial state values
initial_state_values <- c(S = initial_susceptible, E = initial_exposed, I = initial_infected, R = initial_recovered)

# When V =0.0 and sigma =0.0
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.0, mu = leaving_rate, alpha = infected_rate, sigma = 0.0)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot1 <- ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ = 0.0 & V = 0.0",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# When V =0.0 and sigma =0.25
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.0, mu = leaving_rate, alpha = infected_rate, sigma = 0.25)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot2 <-ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ = 0.25 & v = 0.0",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# Arrange plots side by side
print(plot1)
print(plot2)
grid.arrange(plot1, plot2, ncol = 2)
```
### EXPLANATION
If none of the exposed individuals at the latent period are diagnosed and treatedσ, and there is no control measure been introduced into the susceptible class,(v). It will take a much longer time for exposed individuals to decrease, longer period before we notice any significant improvement for individuals to recovered from the measles disease and the infected individuals will significantly increase before noticing a drop in the number of infected individuals. 
On the hand when we increase the exposed individuals at the latent period are diagnosed and treated,σ, by 25% there is a clear improvement in the result.

```{r}
# When V =0.25 and sigma =0.25
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.25, mu = leaving_rate, alpha = infected_rate, sigma = 0.25)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot3 <- ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ = 0.25 & v = 0.25",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# When σ =0.50 and v =0.25
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.25, mu = leaving_rate, alpha = infected_rate, sigma = 0.50)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot4 <-ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ =0.50 & v =0.25",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# Arrange plots side by side
print(plot3)
print(plot4)
grid.arrange(plot3, plot4, ncol = 2)
```

### EXPLANATION
If 25% of susceptible individuals are vaccinated,(v), in addition to increases of 25% to 50% of the exposed individuals at the latent period who are diagnosed and treated(σ). It will take lesser time for exposed individuals to decrease significantly,lesser time before noticing
any significant improvement for individuals to recovered from the disease and the infected individuals will significantly decline over time

```{r}
# When σ =0.50 and v =0.50
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.50, mu = leaving_rate, alpha = infected_rate, sigma = 0.50)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot5 <-ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ =0.50 & v =0.50",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# When σ =0.75 and v =0.50
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.75, mu = leaving_rate, alpha = infected_rate, sigma = 0.50)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot6 <-ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ=0.75 & v=0.50",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# Arrange plots side by side
print(plot5)
print(plot6)
grid.arrange(plot5, plot6, ncol = 2)
```
### EXPLANATION
If 50% of susceptible individuals are vaccinated,(v), in addition to increases of 50% to 75% of the exposed individuals at the latent period, who are diagnosed and treated(σ). There is a sudden decline of exposed individuals with time, it takes a  much lesser period for individuals to recover from the disease and lastly infected individuals will take a lesser period to go down over time.

```{r}
# When σ =0.50 and v =0.75
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.70, mu = leaving_rate, alpha = infected_rate, sigma = 0.50)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot7 <- ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ=0.50 & v =0.75",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# When σ =0.75 and v =0.75
# Parameters
parameters <- c(beta = exposure_rate, gamma = recovery_rate, N = population_size, b = recruitment_rate, v = 0.75, mu = leaving_rate, alpha = infected_rate, sigma = 0.75)

# Time vector with increased resolution
time <- seq(0, simulation_period, by = 0.1)  

# Solve the model with the updated time vector
output <- ode(y = initial_state_values, times = time, func = seir_model, parms = parameters)

# Convert the output to a data frame
output_df <- as.data.frame(output)
colnames(output_df) <- c("time", "Susceptible", "Exposed", "Infected", "Recovered")

# Convert the data to a tidy format
dat <- output_df %>% pivot_longer(cols = c(Susceptible, Exposed, Infected, Recovered), names_to = "state", values_to = "value")

# Plot the SEIR model with the increased resolution
plot8 <-ggplot(data = dat, aes(x = time, y = value, color = state)) +
  geom_line() +
  labs(title = "SEIR Model,σ=0.75 & v =0.75",
       x = "Time",
       y = "Number of Individuals") +
  scale_color_manual(values = c("Susceptible" = "blue",
                                 "Exposed" = "purple",
                                 "Infected" = "red",
                                 "Recovered" = "green")) +
  theme_minimal()

# Arrange plots side by side
print(plot7)
print(plot8)
grid.arrange(plot7, plot8, ncol = 2)
```
### EXPLANATION
If 75% of exposed individuals at the latent period are
diagnosed and treated(σ). In addition to increases of 50% to 75% of susceptible individuals who are vaccinated(v),we will have a better result in comparison to the previous results stated before.

```{r}
# Arrange plots side by side with larger size
grid.arrange(plot1, plot2, plot3, plot4, ncol = 2, widths = c(8, 8),heights = c(5, 5))
grid.arrange(plot5, plot6, plot7, plot8, ncol = 2, widths = c(8, 8),heights = c(5, 5))
```
Impact of Parameters (v and σ):

v: Represents the rate at which individuals become immune to the disease through vaccination or other means.
σ: Represents the rate at which exposed individuals are identified and receive treatment.
Simulation Results:

We can conclude that an increase in either v and σ will results to fewer infected, exposed and lesser period before noticing any significant improvement for individuals to recovered from the disease. There will be a faster recovery rate.

### Application of R0 to measles in Nigeria
```{r}
# Parameter 
b <- 0.03691  
v <- c(0.0, 0.25, 0.50, 0.75) 
mu <- 0.01241  
beta <- 0.09091  # per day
gamma <- 0.125  # per day
sigma <- c(0.0, 0.25, 0.50, 0.75)  
alpha <- 0.14286  # per day

# Calculate R0 for each value of σ
R0_values <- alpha * beta * b / (mu * (mu + alpha + sigma) * (mu + gamma))
R0_values

# Create data frame for plotting
df_sigma <- data.frame(sigma = sigma, R0 = R0_values)

# Plot R0 values against sigma
library(ggplot2)
plot_sigma <- ggplot(df_sigma, aes(x = sigma, y = R0)) +
  geom_point(size = 3) +
  geom_line(method = "loess", se = FALSE, color = "red") +  # Add line
  labs(title = "Effect of sigma on Reproductive Number (R0)",
       x = "sigma (Exposed Individuals Testing Rate)",
       y = "R0 (Basic Reproduction Number)") +
  theme_minimal()

# Print the plot
print(plot_sigma)

```
### EXPLANATION

The Basic Reproduction Number, R0, is a key measure in understanding disease spread. It tells us how many new cases one infected person will cause in a susceptible population. If R0 is above 1, the disease will likely spread, while if it's below 1, the disease will eventually die out. In our model, R0 changes with the rate of exposed individuals undergoing testing and therapy, denoted as σ.

When σ is 0, R0 is above 1, indicating that without testing and therapy, the disease will persist in the population. As σ increases up to 1, R0 decreases, suggesting better control and potential elimination of the disease. This reflects how effective testing and therapy strategies impact disease spread.

In mathematical modelling stability refers to whether the disease is likely to persist within a population or die out over time.
Regarding stability, when R0 is less than 1, it means the disease-free state is stable, and the population is likely to remain disease-free over time. Conversely, if R0 is above 1, the disease-free state is unstable, and the disease will persist within the population, potentially spreading further. This understanding of R0 acts as a threshold for determining whether the disease will establish itself or die out in a population.